Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-10-23 causal-bandit/nips_2016/additional_experiments.tex
updates additional experiments and proofs

In this section we compare performance of our algorithms with the standard bandit algorithms on the confounded graph in figure \ref{fig:parallel_confounded}. All the variables are binary and the action space, $\calA$, consists of the set of single variable interventions $do(V_i = j)$. We choose this setting as it can be seen as a generalisation of the parallel bandit case and allows for factorizations that enable us to compute the exact reward distribution and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$). As before, we show the average regret over 10,000 simulations with error bars showing three standard errors.

@@

As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1)=1) = \frac{1}{2} + \epsilon$ amd $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon\frac{P(X_1=1)}{P(X_1=0)}$. We vary $N_1$ from 2 to $N$ to vary $m$. The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$. This allows us to just show the dependence on $m$, removing the noise associated with different models producing selecting values for $\eta*$ with the same $m$ (and thus same worst case performance) but different performance for a given reward distribution.

@@

In figure \ref{fig:simple_vs_T_misspecified} we exclude actions on $Z$ from the set of allowable actions to demonstrate how algorithm 1 can fail in the presence of a hidden confounding variable. Algorithm 1 incorrectly assumes $P(Y|do(X)) = P(Y|X)$ leading to biased estimates. We let $P(Z) = .52$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:

@@

The first group of $X$ variables with $i \leq 6$ will be identified by the parallel bandit as the most unbalanced ones and played explicitly. All remaining variables are likely to be identified as balanced and estimated from observational estimates. This leads to the worst case outcome, where the bias in the estimates leads Algorithm 1 to asymptotically select a sub-optimal action.

\begin{figure}[H]
    \begin{subfigure}[t]{0.3\textwidth}
		\centering    
    		\includegraphics[width=\textwidth]{experiment4_20161020_1254.pdf}
    		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
        \label{fig:simple_vs_m_general}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
        \includegraphics[width=\textwidth]{experiment7_20161020_1257.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$ and $m(\eta*)=3.1$ }
        \label{fig:simple_vs_T_general}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
    		\includegraphics[width=\textwidth]{experiment5_20161020_2027.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 20$, $m=4.3$ with no actions setting $Z$}
    		\label{fig:simple_vs_T_misspecified}
    \end{subfigure}
    \caption{Experimental results on the confounded graph}
    \label{fig:experiments}
\end{figure}

>>>>

In this section we compare performance of our algorithms with the standard bandit algorithms on the confounded graph in figure \ref{fig:parallel_confounded}. All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action, $\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}$. We choose this setting as it can be seen as a generalisation of the parallel bandit case and allows for factorizations that enable us to compute the exact reward distribution and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$). As before, we show the average regret over 10,000 simulations with error bars showing three standard errors.

@@

As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ amd $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon\frac{P(X_1=1)}{P(X_1=0)}$. We vary $N_1$ from 2 to $N$ to vary $m$. The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$. This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta*$ with the same $m$ (and thus same worst case performance) but different performance for a given reward distribution.

@@

In figure \ref{fig:simple_vs_T_misspecified} we exclude actions on $Z$ from the set of allowable actions to demonstrate how algorithm 1 can fail in the presence of a confounding variable. Algorithm 1 incorrectly assumes $P(Y|do(X)) = P(Y|X)$ leading to biased estimates. We let $P(Z) = .6$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:

@@

In this setting $X_7$ tends to agree with $Z$ and $X_N$ tends to disagree. It is sub-optimal to act on either $X_7$ or $X_N$, all other actions are optimal. The first group of $X$ variables with $i \leq 6$ will be identified by the parallel bandit as the most unbalanced ones and played explicitly. All remaining variables are likely to be identified as balanced and estimated from observational estimates. The CPD values have been chosen to demonstrate the worst case outcome, where the bias in the estimates leads Algorithm 1 to asymptotically select a sub-optimal action.

\begin{figure}[H]
    \begin{subfigure}[t]{0.3\textwidth}
		\centering    
    		\includegraphics[width=\textwidth]{experiment4_20161020_1254.pdf}
    		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
        \label{fig:simple_vs_m_general}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
        \includegraphics[width=\textwidth]{experiment7_20161020_1257.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$ and $m(\eta*)=3.1$ }
        \label{fig:simple_vs_T_general}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
    		\includegraphics[width=\textwidth]{experiment5_20161021_2026.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 21$, $m=4.3$ with no actions setting $Z$}
    		\label{fig:simple_vs_T_misspecified}
    \end{subfigure}
    \caption{Experimental results on the confounded graph}
    \label{fig:experiments_confounded}
\end{figure}

### 2016-10-23 causal-bandit/nips_2016/tech.tex
updates additional experiments and proofs

\begin{lemma}\label{lem:conc2}
Let $X_1,X_2\ldots,$ be a sequence of random variables with $X_i \in [0,1]$ and $\EE[X_i] = p$ and $\delta \in [0,1]$. 
\eq{
\P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} \leq 4\delta\,.
}
\end{lemma}

\begin{proof}
For $\delta \geq 1/4$ the result is trivial. Otherwise, Hoeffding's inequality gives,
\eq{
\P{\left|\frac{1}{t}\sum_{s=1}^t{X_s} - p \right| > \epsilon} \leq 2 e^{-2t\epsilon^2}
} Let $\epsilon = \sqrt{\frac{2}{n_0}\log{\frac{2}{\delta}}}$ and taking the union bound over $t\geq n_0$,

\eq{
\P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} 
&\leq 2\sum_{t=n_0}^\infty \exp\left(-\frac{4t}{n_0} \log \frac{2}{\delta}\right)\\
& = 2\frac{r^{n0}}{1-r} \text{ where } r = (\frac{\delta}{2})^{4/n_0}
}
\end{proof}
THIS WOULD SEEM TO DIVERGE AS $n_0$ goes to $\infty$. HOW DO WE GET $\leq 4\delta\,. \qedhere$

@@

\begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then
for $a \in A$ the algorithm estimates $\mu_a$ from $T/(2m(\vec{\hat q})) \geq T/(4m(\vec{q}))$ samples.
Therefore by Hoeffding's inequality and the union bound we have
\eq{
\P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}} \leq \delta\,.
}
For arms not in $A$ we have $\hat p_a \geq 1/m(\vec{\hat q})$ For $a = do(X_i = j)$, if $a = do(X_i = j)$, then from Lemma \ref{lem:m_est},
\eq{
\hat p_a = \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{1}{2m(\vec{q})}\, \text{ with probability } 1-\delta
}

Let $H$ be the event $\sum_{t=1}^{T/2} \ind{X_{t,i} = j} < T/4m(\vec{q})$ and $G$ be the event $\exists a \notin A:  \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{8m}{T} \log \frac{2N}{\delta}}$.

\eq{
P(G) & \leq P(H) + P(G|\neg H)\\
& \leq \delta + 4\delta \longleftarrow \text{ by Lemma \ref{lem:conc2}, letting } n_0 = \frac{T}{4m(\vec{q})}
}

WHY DO WE NEED THE UNION BOUND AND THUS Lemma \ref{lem:conc2}. For each of these arms we select them some $t > n_0$ times. This $t$ is a random variable but doesn't depend on the empirical reward estimates for $s \in {n_0 ... t}$.

Therefore, combining this result with the bound for arms $a \in A$, we have with probability at least $1 - 6\delta$ that,
\eq{
(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{8m}{T} \log \frac{2N}{\delta}} = \epsilon\,.
}
If this occurs, then 
\eq{
\mu_{\hat a^*_T} \geq \hat \mu_{\hat a^*_T} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
}
Therefore
\eq{
\mu^* - \EE[\mu_{\hat a^*_T}] 
& \leq 6\delta + \epsilon\\
& \leq \frac{6m}{T} + \sqrt{\frac{16m}{T} \log \frac{NT}{m}}, \longleftarrow \text{ letting } \delta = \frac{m}{T}\\
& \leq \sqrt{\frac{44m}{T} \log \frac{NT}{m}}\ \longleftarrow \text{ via Jenson's Inequality}
}
which completes the result.
\end{proof}

>>>>

@@

\begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then,
for $a \in A$, the algorithm estimates $\mu_a$ from $T_A \doteq T/(2m(\vec{\hat q}))$ samples. From lemma \ref{lem:m_est}, $T_A  \geq T/(4m(\vec{q}))$ with probability $(1-\delta)$. Let $H$ be the event $T_A < T/(4m(\vec{q}))$ and $G$ be the event $\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}$

\eq{
\P{G} \leq \P{H} + \P{G|\neg H} \leq \delta + \P{G|\neg H}
}

Via Hoeffding's inequality and the union bound,

\eq{
\P{G|\neg H}&\doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}},\text{ given }T_A \geq T/(4m(\vec{q}))} \leq \delta\\
 \implies  \P{G} & \doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}} \leq 2\delta\,.
}
For arms not in $A$,

\eq{
\hat p_a = & \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq 1/m(\vec{\hat q}), \; \text {by definition of not being in $A$}\\
 \geq & \frac{1}{2m(\vec{q})},\; \text{ with probability } 1-\delta \\
\implies  T_a \doteq & \sum_{t=1}^{T/2} \ind{X_i = j} \geq  \frac{T}{4m(\vec{q})},\; \text{ with probability } 1-\delta
}

Again applying Hoeffding's and the union bound

\eq{
\P{\exists a \notin A:  \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}}} \leq 2\delta
}

Therefore, combining this result with the bound for arms $a \in A$, we have with probability at least $1 - 4\delta$ that,
\eq{
(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}} \doteq \epsilon\,.
}
If this occurs, then 
\eq{
\mu_{\hat a^*_T} \geq \hat \mu_{\hat a^*_T} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
}
Therefore
\eq{
\mu^* - \EE[\mu_{\hat a^*_T}] 
& \leq 4\delta + \epsilon\\
& \leq \frac{8m(\vec{q})}{T} + \sqrt{\frac{2m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ letting } \delta = \frac{2m(\vec{q})}{T}\\
& \leq \sqrt{\frac{20m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ via Jenson's Inequality}
}
which completes the result.
\end{proof}
